Block bootstrap consistency\\under weak assumptions

### 2015-12-29 Paper.tex

\documentclass[11pt]{article}
\usepackage{amsfonts}
\usepackage{amsmath,amsthm,amssymb,graphicx,setspace,url,booktabs,tabularx,slantsc}
\usepackage[T1]{fontenc}

% \usepackage[yyyymmdd,hhmmss]{datetime}
% \usepackage{fancyhdr}
% \pagestyle{fancy}
% \lfoot{Page \thepage. Typeset on \today\ at \currenttime}
% \cfoot{}
% \rfoot{}

\usepackage[sort,round]{natbib}
\usepackage[multiple]{footmisc}
\usepackage{geometry}
\usepackage[small]{caption}

\bibliographystyle{abbrvnat}
\newcommand\citepos[2][]{\citeauthor{#2}'s \citeyearpar[#1]{#2}}

\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{cor}{Corollary}
\newtheorem{res}{Result}

\theoremstyle{definition}

\newtheorem{example}{Example}
\newtheorem{defn}{Definition}
\newtheorem{rem}{Remark}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\plim}{plim}

\DeclareMathOperator{\dist}{d}
\DeclareMathOperator{\const}{const}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\eqd}{\overset{d}{=}}
\DeclareMathOperator{\ind}{1}
\DeclareMathOperator{\pr}{Pr}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\vech}{vech}

\renewcommand{\mod}{\operatorname{mod}}

\newcommand{\clt}{\textsc{clt}}
\newcommand{\fclt}{\textsc{fclt}}
\newcommand{\hac}{\textsc{hac}}
\newcommand{\lln}{\textsc{lln}}
\newcommand{\mds}{\textsc{mds}}
\newcommand{\ned}{\textsc{ned}}

\frenchspacing

@@

\author{Gray Calhoun\thanks{Economics Department, Iowa State
    University, Ames, IA 50011. Telephone: (515) 294-6271.  Email:
    \texttt{gcalhoun@iastate.edu}. Web: \texttt{http://gray.clhn.co}.
    I would like to thank Helle Bunzel, Dimitris Politis, Robert
    Taylor, and three anonymous referees for their comments and
    feedback on earlier versions of this paper.}}

@@

\date{March 25, 2013}
\maketitle

\begin{abstract}\noindent
  This paper weakens the size and moment conditions needed for typical
  block bootstrap methods (i.e. the Moving Blocks, Circular Blocks,
  and Stationary Bootstraps) to be valid for the sample mean of
  Near-Epoch-Dependent (\ned) functions of mixing processes; they are
  consistent under the weakest conditions that ensure the original
  \ned\ process obeys a Central Limit Theorem \citep[those
    of][\textit{Econometric Theory}]{Jon:97}.  In doing so, this paper
  extends de Jong's method of proof, a blocking argument, to hold with
  random and unequal block lengths.  This paper also proves that
  bootstrapped partial sums satisfy a Functional CLT under the same
  conditions.

@@

\newpage
\noindent Block bootstraps, e.g. the Moving Blocks
\citep{Kun:89,LiS:92}, Circular Block \citep{PoR:92}, and Stationary
Bootstraps \citep{PoR:94}, have become popular in Economics, partly
because they do not require the researcher to make parametric
assumptions about the data generating process.  They are valid under
general weak dependence and moment conditions.  Some recent papers
(\citealp{GoW:02}; \citealp{GoJ:03}) relax the dependence and moment
conditions of the original papers to fit with those commonly used in
Econometrics based on Near-Epoch-Dependence
(\ned).\footnote{\citet{GoW:02} show that these bootstrap methods can
  be applied to heterogeneous $L_{2+\delta}$-\ned\ processes of size
  $-2(r-1)/(r-2)$ on a strong mixing sequence of size
  $-r(2+\delta)/(r-2)$, where $r > 2$ and $\delta >0$, when the
  original series has uniformly bounded $3r$-moments.  \citet{GoJ:03}
  relax these conditions to $L_{2+\delta}$-\ned\ of size $-1$ and
  $r+\delta$ moments for the original series, and size
  $-(2+\delta)(r+\delta)/(r-2)$ for the underlying mixing series.
  Both papers require that the expected block length grow with $n$ and
  be $o(n^{1/2})$.  \cite{GoP:11} discuss these issues
  further.}\footnote{An array $\{X_{nt}\}$ is an $L_{\rho}$-\ned\
  process on a mixing array $\{V_{nt}\}$ if
  \begin{equation}
    \| X_{nt} - \E(X_{nt}
    \mid V_{n,t-m},\dots,V_{n,t+m}) \|_{\rho} \leq d_{nt} v_m
  \end{equation} 
  with $v_m \to 0$ as $m \to \infty$ and $\{d_{nt}\}$ an array of
  positive constants.  It is of size $-\gamma$ if $v_m = O(m^{-\gamma
    - \delta})$ for all $\delta>0$.  Dropping the index ``$n$'' gives
  the series definition.  Note that strong and uniform mixing arrays are
  not required to be stationary.} But these conditions are still stronger than
required for a \clt\ to hold; \citet{Jon:97} has established the \clt\
under $L_2$-\ned\ with smaller size and moment
restrictions.\footnote{\citet{Jon:97} proves that the \clt\ holds for
  averages of $L_2$-\ned\ processes of size $-1/2$ on a strong mixing
  series of size $-r/(r-2)$, $r > 2$ and the original series having
  bounded $r$-moments.}  This paper shows that these block bootstrap
methods consistently estimate the distribution of the sample mean
under \citepos{Jon:97} assumptions, and show that an \fclt\ holds as
well.\footnote{\citet{Rad:96} proves consistency for the Moving Blocks
  Bootstrap for any stationary strong mixing sequence that satisfies
  the \clt. This paper uses a similar method of proof to his, but also
  accommodates nonstationary sequences and the Stationary Bootstrap.}
It also relaxes \citepos{GoW:02} and \citepos{GoJ:03} requirement that
the expected block length be $o(n^{1/2})$ to the original papers'
requirement that it be $o(n)$.

The proof exploits the conditional independence of the blocks in each
bootstrap.  Each bootstrap proceeds by drawing blocks of $M$
consecutive observations from the original time series, and then
pasting these blocks together to create the new bootstrap time series.
The Moving Blocks bootstrap does exactly that; the Circular Block
bootstrap ``wraps'' the observations, so that $(X_{n-1}, X_n, X_1,
X_2)$, for example, is a possible block of length four (letting $X_t$
denote the original time series).  The Stationary Bootstrap wraps the
observations and also draws $M$ at random for each block;
\citet{PoR:94} suggest drawing $M$ from the geometric distribution.
As the name suggests, the series produced by the Stationary Bootstrap
are strictly stationary, while those produced by the other methods are
not.  Although the Stationary Bootstrap was believed to be much less
efficient than other block bootstrap methods due to results of
\citet{Lah:99}, \citet{Nor:09} has shown that it is only slightly less
efficient than the other block-bootstrap methods discussed in this
paper, and has efficiency identical to that of the non-overlapping
block bootstrap.  Consequently, there has been renewed interest in the
Stationary Bootstrap since stationarity of the bootstrap samples can
be a useful property in theoretical research.  \cite{KrP:11} provides
a recent review of the bootstrap for time-series
processes,\footnote{Also see the discussion papers:
  \citet{Dah:11,GoP:11,Hor:11,JeM:11,KrP:11b}.} and \citet{GoP:11}
further discuss recent developments in block-bootstraps.

Theorem~\ref{thm:1} presents the main result, asymptotic normality of
the distribution of bootstrapped partial sums.  This paper adopts the
standard notation that $\E^{*}$, $\var^{*}$, etc. are the usual
operators with respect to the probability measure induced by the
bootstrap and will use explicit stochastic array notation 
for precision.  Also note that all results are presented for the
scalar case but generalize immediately to random vectors.  All of the
proofs are presented in the appendix; only proofs for the Stationary
Bootstrap are presented, since proofs for the other methods are
similar and easier to follow.  All limits are taken as $n \to \infty$
unless otherwise noted and $\lVert \cdot \rVert_r$ denotes the
$L_r$-norm.

\begin{thm}\label{thm:1}
  Suppose the following conditions hold.
  \begin{enumerate}
  \item $X_{nt}$ is $L_2$-\ned\ of size $-1/2$ on an array
    $V_{nt}$ that is either strong mixing of size $-r/(r-2)$ or
    uniform mixing of size $-r/2(r-1)$, with $r > 2$.  The
    \ned\ magnitude indices are denoted $\{d_{nt}\}$.
  \item The array $\mu_{nt} - \bar \mu_n$ is uniformly bounded where
    $\E X_{nt} = \mu_{nt}$ and $\bar{\mu}_n = n^{-1} \sum_{t=1}^n
    \mu_{nt}$.  Moreover, $\sqrt{n} \| \bar{X}_{n} - \bar\mu_n \|_2
    \to \sigma > 0$.
  \item There exists an array of positive real numbers $\{c_{nt}\}$
    such that $(X_{nt} - \mu_{nt})/c_{nt}$ is uniformly $L_r$-bounded
    and $c_{nt}$ and $d_{nt}/c_{nt}$ are uniformly bounded in $n$
    and~$t$.
  \item $X_{nt}^{*}$ is generated by the Stationary Bootstrap with
    geometric block lengths with success probability $p_n$, $p_n = c
    n^{-a}$ and $a,c \in (0,1)$, or by the Moving or Circular Block
    bootstrap with block length $M_n$ such that $M_n \sim n^a$ for 
    $a \in (0,1)$.  Let $M_{ni}$ be the block length of the $i$th
    block, $i=1,\dots,J_n$, and define $K_{n0} = 0$ and $K_{nj} =
    \sum_{i=1}^j M_{ni}$.
  \end{enumerate}
  Then
  \begin{equation}\label{eq:30}
    \sup_x \big\lvert \pr^{*}\big[\sqrt{n}(\bar X_{n}^{*} - \mu_{n}^{*}) \big/
    \sigma_n^{*} \leq x \big] - \Phi(x) \big\rvert \to^p 0,
  \end{equation}
  and
  \begin{equation}
    \label{eq:9}
    \sup_x \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} - \mu_{n}^{*}) \big/ \hat\sigma_n^{*}
    \leq x \big] - \Phi(x) \big\rvert \to^p 0,
  \end{equation}
  where $\Phi$ is the \textsc{cdf} of the Standard Normal
  distribution, $\sigma_n^{*2} = \var^{*}(\sqrt{n} \bar X_n^{*})$, and
    \begin{equation}
      \label{eq:12}
      \hat{\sigma}_n^{*2} = \tfrac{1}{n} \sum_{j=1}^{J_n}
      \Big\{\sum_{t=K_{n,j-1}+1}^{K_{nj}} (X_{nt}^{*} - \bar X_{n}^{*})\Big\}^2.
    \end{equation}
\end{thm}

Note that \citet{Jon:97} allows a little bit more flexibility in his
conditions on the array $\{c_{nt}\}$ \citep[see also][]{Dav:93};
essentially, he allows for a single set of blocks with the
maximal $\{c_{nt}\}$ over each block well-behaved, while this
paper requires this condition to hold for every possible partition of
blocks.  This additional restriction is required because the
Stationary Bootstrap will select the blocks randomly and is similar to
\citepos{JoD:00b} for the \fclt.

Consistency for the distribution of the sample mean follows as an
immediate corollary of Theorem~\ref{thm:1} and \citet[Theorem~2]{Jon:97}.

\begin{cor}\label{cor:1} 
  If the conditions of Theorem~\ref{thm:1} hold and $\hat{\sigma}^2_n$
  is a consistent estimator of $\sigma^2$ then
  \begin{equation}
    \label{eq:31}
    \sup_{x} \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} - \mu_{n}^{*}) / \sigma_{n}^{*}
    \leq x \big]
    - \pr\big[\sqrt{n}(\bar X_n - \bar{\mu}_n)/\hat{\sigma}_n \leq x \big] \big\rvert \to^p 0.
  \end{equation}
  and
  \begin{equation}
    \label{eq:11}
    \sup_{x} \big\lvert \pr^{*}\big[
    \sqrt{n} (\bar X_{n}^{*} - \mu_{n}^{*}) \big/ \hat\sigma^{*}_{n}
    \leq x \big]
    - \pr\big[\sqrt{n}(\bar X_n - \bar{\mu}_n)/\hat{\sigma}_n \leq x \big] \big\rvert \to^p 0.
  \end{equation}
\end{cor}

Corollary~\ref{cor:1} justifies using the Stationary Bootstrap to
conduct inference about $\bar \mu_n$ even though there may be
considerable heterogeneity.  Unlike \citet{GoW:02} and \citet{GoJ:03},
we are able to achieve consistency without
additional assumptions that ensure the heterogeneity is asymptotically
irrelevant, either because the deviations of $\mu_{nt}$ from $\bar
\mu_n$ are small or because there are a finite number of breaks that
occur in a neighborhood of the first observation.
The additional cost is the assumption that there must be an estimator
$\hat\sigma^2$ of $\sigma^2$ that is known to be consistent, and such an
estimator will typically require the heterogeneity in the mean to be
approximately known.

If the mean \textit{is} approximately constant, we can get a result
directly comparable to \citepos{GoW:02} and \citet{GoJ:03}, but with
the weaker size and moment conditions of Theorem~\ref{thm:1}.
Corollary~\ref{cor:2} formalizes this result and follows from
Corollary~\ref{cor:1} and Lemma~\ref{res:B}.
\begin{cor}\label{cor:2}
  Suppose that the conditions of Theorem~\ref{thm:1} hold and that
  $\sum_{t=1}^n (\mu_{nt} - \bar{\mu}_n)^2$ is $o(n^{1/2})$.  
  Then $\sigma_n^{*2} \to^p
  \sigma^2$, $\hat{\sigma}_n^{*2} \to^p \sigma^2$, and
\end{cor}
  \begin{equation}
    \sup_{x} \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} - \mu_{n}^{*})
    \leq x \big]
    - \pr\big[\sqrt{n}(\bar X_n - \bar{\mu}_n) \leq x \big] \big\rvert \to^p 0.
  \end{equation}

For some uses, a \clt\ is not enough and an \fclt\ is desired.
Theorem~\ref{thm:1} can be extended to give an \fclt\ (using
arguments from \citealp{JoD:00b}), but we will see that extending 
Corollaries~\ref{cor:1} and~\ref{cor:2} is more problematic.
Theorem~\ref{thm:2} proves a version of Theorem~\ref{thm:1} for the
partial sum; under the same assumptions as Theorem~\ref{thm:1}, the
partial sum of the bootstrapped process obeys an \fclt\ and can be
used to derive critical values for other test statistics.

\begin{thm}\label{thm:2}
  Suppose that the conditions of Theorem~\ref{thm:1} hold, let $W$ be
  standard Brownian Motion, and define
  \begin{equation}
    \label{eq:7}
    W_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \sigma_{n}^{*}
  \end{equation}
  and 
  \begin{equation}
    \label{eq:7}
    \hat{W}_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \hat\sigma_n^{*}.
  \end{equation}
  Then $\pr^{*}[\dist(W_n^{*}, W) > \delta] \to 0$ i.p. and
  $\pr^{*}[\dist(\hat{W}_n^{*}, W) > \delta] \to 0$ i.p. for any
  positive $\delta$ and distance $\dist$ that metricizes weak
  convergence.
\end{thm}

Just as with Theorem~\ref{thm:1}, we have natural corollaries using
Theorem~\ref{thm:2} to approximate the behavior of partial sums of the
original series.  But applying this corollary requires much stronger
assumptions.  In Corollary~\ref{cor:1}, the
heterogeneity in $X_{nt}$ affects inference through the estimator of
the asymptotic variance, $\hat{\sigma}_n^2$; but if the researcher
has a consistent estimator of the variance, inference on the grand
mean, $\bar{\mu}_{n}$, can proceed as normal.
Assuming negligible heterogeneity in the means was unnecessary unless
we want to avoid estimating $\sigma^2$ directly.

But for the partial sum, a consistent estimator of the variance is not
enough.  If the mean is not approximately constant then only the full
summation is centered at the grand mean---the partial sums are
centered at partial averages of the $\mu_{nt}$ terms.  If the research
question concerns the individual $\mu_{nt}$'s then the bootstrap
approximation can be used, but otherwise it can not.

A second difference is that, if the $\mu_{nt}$ terms are approximately
constant as in Corollary~\ref{cor:2}, the covariance process must
still be consistently estimated.  The bootstrapped processes are
(essentially) homoskedastic, so they are unable to match patterns of
heteroskedasticity in the original series.  Those patterns must be
removed by studentizing both the original and bootstrap partial sums.
Other methods, such
as the Local Block Bootstrap \citep{PaP:02,DPP:03}, may be able to
capture this additional heterogeneity, but we do not pursue that
possibility further.

\begin{cor}
  Suppose the conditions of Theorem~\ref{thm:1} hold, define
  \begin{equation}
    \label{eq:14}
    \sigma_n^2(\gamma) = \tfrac{1}{n}
    \sum_{s,t=1}^{\lfloor \gamma n \rfloor} \cov(X_{ns}, X_{nt}),
  \end{equation}
  let $\hat\sigma_n^2(\cdot)$ be an estimator of $\sigma_n^2(\cdot)$
  such that $\sup_{\gamma} |\sigma_n^2(\gamma) /
  \hat{\sigma}_n^2(\gamma) - 1| \to^p 0$, and let
  \begin{equation}
    \label{eq:8}
    \hat{W}_n(\gamma) = \tfrac{1}{\sqrt{n}} \sum_{t=1}^{\lfloor \gamma
      n \rfloor} (X_{nt} - \mu_{nt}) / \hat{\sigma}_{n}(\gamma).
  \end{equation}
  Then $\pr^{*}[\dist(W_n^{*}, \hat{W}_n) > \delta] \to^p 0$ and
  $\pr^{*}[\dist(\hat{W}_n^{*}, \hat{W}_n) > \delta] \to^p 0$ for any
  positive $\delta$ and any distance $\dist$ that metricizes weak
  convergence.

>>>>

\documentclass[11pt]{article}
\input{preamble}

@@

\author{Gray Calhoun\thanks{Economics Department, Iowa State
    University, Ames, IA 50011. Telephone: (515) 294-6271.  Email:
    \protect\url{gcalhoun@iastate.edu}. Web:
    http://www.econ.iastate.edu/$\sim$gcalhoun.  I would like to thank
    Helle Bunzel, Dimitris Politis, Robert Taylor, and three anonymous
    referees for their comments and feedback on earlier versions of
    this paper.}}

@@

\maketitle

\begin{abstract}\noindent
  This paper weakens the size and moment conditions needed for typical
  block bootstrap methods (i.e.\ the Moving Blocks, Circular Blocks,
  and Stationary Bootstraps) to be valid for the sample mean of
  Near-Epoch-Dependent (\ned) functions of mixing processes; they are
  consistent under the weakest conditions that ensure the original
  \ned\ process obeys a Central Limit Theorem \citep[those
    of][\textit{Econometric Theory}]{Jon:97}.  In doing so, this paper
  extends de Jong's method of proof, a blocking argument, to hold with
  random and unequal block lengths.  This paper also proves that
  bootstrapped partial sums satisfy a Functional \clt\ under the same
  conditions.

@@

\newpage
\noindent Block bootstraps, e.g.\ the Moving Blocks
\citep{Kun:89,LiS:92}, Circular Block \citep{PoR:92}, and Stationary
Bootstraps \citep{PoR:94}, have become popular in Economics, partly
because they do not require the researcher to make parametric
assumptions about the data generating process.  They are valid under
general weak dependence and moment conditions.  Some recent papers
(\citealp{GoW:02}; \citealp{GoJ:03}) relax the dependence and moment
conditions of the original papers to fit with those commonly used in
Econometrics based on Near-Epoch-Dependence (\ned).%
\footnote{\citet{GoW:02} show that these bootstrap methods can be
  applied to heterogeneous $L_{2+\delta}$-\ned\ processes of size
  $-2(r-1)/(r-2)$ on a strong mixing sequence of size
  $-r(2+\delta)/(r-2)$, where $r > 2$ and $\delta >0$, when the
  original series has uniformly bounded $3r$-moments.  \citet{GoJ:03}
  relax these conditions to $L_{2+\delta}$-\ned\ of size $-1$ and
  $r+\delta$ moments for the original series, and size
  $-(2+\delta)(r+\delta)/(r-2)$ for the underlying mixing series.
  Both papers require that the expected block length grow with $n$ and
  be $o(n^{1/2})$.  \cite{GoP:11} discuss these issues further.}%
\footnote{An array $\{X_{nt}\}$ is an $L_{\rho}$-\ned\ process on a
  mixing array $\{V_{nt}\}$ if
  \begin{equation}
    \| X_{nt} - \E(X_{nt}
    \mid V_{n,t-m},\dots,V_{n,t+m}) \|_{\rho} \leq d_{nt} v_m
  \end{equation}
  with $v_m \to 0$ as $m \to \infty$ and $\{d_{nt}\}$ an array of
  positive constants.  It is of size $-\gamma$ if $v_m = O(m^{-\gamma
    - \delta})$ for all $\delta>0$.  Dropping the index ``$n$'' gives
  the series definition.  Note that strong and uniform mixing arrays
  are not required to be stationary.} %
But these conditions are still stronger than
required for a \clt\ to hold; \citet{Jon:97} has established the \clt\
under $L_2$-\ned\ with smaller size and moment
restrictions.%
\footnote{\citet{Jon:97} proves that the \clt\ holds for averages of
  $L_2$-\ned\ processes of size $-1/2$ on a strong mixing series of
  size $-r/(r-2)$, $r > 2$ and the original series having bounded
  $r$-moments.} %
This paper shows that these block bootstrap
methods consistently estimate the distribution of the sample mean
under \citepos{Jon:97} assumptions, and show that an \fclt\ holds as
well.%
\footnote{\citet{Rad:96} proves consistency for the Moving Blocks
  Bootstrap for any stationary strong mixing sequence that satisfies
  the \clt. This paper uses a similar method of proof to his, but also
  accommodates nonstationary sequences and the Stationary
  Bootstrap.} %
It also relaxes \citepos{GoW:02} and \citepos{GoJ:03} requirement that
the expected block length be $o(n^{1/2})$ to the original papers'
requirement that it be $o(n)$.

The proof exploits the conditional independence of the blocks in each
bootstrap.  Each bootstrap proceeds by drawing blocks of $M$
consecutive observations from the original time series, and then
pasting these blocks together to create the new bootstrap time series.
The Moving Blocks bootstrap does exactly that; the Circular Block
bootstrap ``wraps'' the observations, so that $(X_{n-1}, X_n, X_1,
X_2)$, for example, is a possible block of length four (letting $X_t$
denote the original time series).  The Stationary Bootstrap wraps the
observations and also draws $M$ at random for each block;
\citet{PoR:94} suggest drawing $M$ from the geometric distribution.
As the name suggests, the series produced by the Stationary Bootstrap
are strictly stationary, while those produced by the other methods are
not.  Although the Stationary Bootstrap was believed to be much less
efficient than other block bootstrap methods due to results of
\citet{Lah:99}, \citet{Nor:09} has shown that it is only slightly less
efficient than the other block-bootstrap methods discussed in this
paper, and has efficiency identical to that of the non-overlapping
block bootstrap.  Consequently, there has been renewed interest in the
Stationary Bootstrap since stationarity of the bootstrap samples can
be a useful property in theoretical research.  \cite{KrP:11} provides
a recent review of the bootstrap for time-series
processes,%
\footnote{Also see the discussion papers:
  \citet{Dah:11,GoP:11,Hor:11,JeM:11,KrP:11b}.} %
and \citet{GoP:11} further discuss recent developments in
block-bootstraps.

Theorem~\ref{main-bootstrap-clt} presents the main result, asymptotic normality of
the distribution of bootstrapped partial sums.  This paper adopts the
standard notation that $\E^{*}$, $\var^{*}$, etc.\ are the usual
operators with respect to the probability measure induced by the
bootstrap and will use explicit stochastic array notation
for precision.  Also note that all results are presented for the
scalar case but generalize immediately to random vectors.  All of the
proofs are presented in the appendix; only proofs for the Stationary
Bootstrap are presented, since proofs for the other methods are
similar and easier to follow.  All limits are taken as $n \to \infty$
unless otherwise noted and $\lVert \cdot \rVert_r$ denotes the
$L_r$-norm.

\begin{thm}\label{main-bootstrap-clt}
  Suppose the following conditions hold.
  \begin{enumerate}
  \item $X_{nt}$ is $L_2$-\ned\ of size $-1/2$ on an array
    $V_{nt}$ that is either strong mixing of size $-r/(r-2)$ or
    uniform mixing of size $-r/2(r-1)$, with $r > 2$.  The
    \ned\ magnitude indices are denoted $\{d_{nt}\}$.
  \item The array $\mu_{nt} - \bar \mu_n$ is uniformly bounded
    and $\sum_{t=1}^n (\mu_{nt} - \bar \mu_n)^2 \to 0$,
    where $\E X_{nt} = \mu_{nt}$ and $\bar{\mu}_n = n^{-1} \sum_{t=1}^n
    \mu_{nt}$. Moreover, $\sqrt{n} \| \bar{X}_{n} - \bar\mu_n \|_2
    \to \sigma > 0$, with $\bar X_n = (1/n) \sum_{t=1}^n X_{nt}$.
  \item There exists an array of positive real numbers $\{c_{nt}\}$
    such that $(X_{nt} - \mu_{nt})/c_{nt}$ is uniformly $L_r$-bounded
    and $c_{nt}$ and $d_{nt}/c_{nt}$ are uniformly bounded in $n$
    and~$t$.
  \item $X_{nt}^{*}$ is generated by the Stationary Bootstrap with
    geometric block lengths with success probability $p_n$, $p_n = c
    n^{-a}$ and $a,c \in (0,1)$, or by the Moving or Circular Block
    bootstrap with block length $M_n$ such that $M_n \sim n^a$ for
    $a \in (0,1)$.  Let $M_{ni}$ be the block length of the $i$th
    block, $i=1,\dots,J_n$, and define $K_{n0} = 0$ and $K_{nj} =
    \sum_{i=1}^j M_{ni}$. The last block, $M_{n,J_n}$, is defined as
    $K_{n,J_n} - K_{n,J_n-1}$, so $K_{n,J_n} = n$ a.s.
  \end{enumerate}
  Then $\hat\sigma^{*2} \to^p \sigma^2$,
  \begin{equation}\label{eq:36}
    \sup_x \big\lvert \pr^{*}\big[\sqrt{n}(\bar X_{n}^{*} - \E^* \bar X_n^*) \leq x \big]
    - \pr \big[\sqrt{n}(\bar X_{n} - \E \bar X_n) \leq x \big] \big\rvert \to^p 0,
  \end{equation}
  and
  \begin{equation}
    \label{eq:2}
    \sup_x \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} -  \E^* \bar X_n^*) \big/ \hat\sigma_n^{*}
    \leq x \big] - \Phi(x) \big\rvert \to^p 0,
  \end{equation}
  where $\Phi$ is the \textsc{cdf} of the Standard Normal
  distribution and
    \begin{equation}
      \label{eq:3}
      \hat{\sigma}_n^{*2} = \tfrac{1}{n} \sum_{j=1}^{J_n}
      \Big\{\sum_{t=K_{n,j-1}+1}^{K_{nj}} (X_{nt}^{*} - \bar X_{n}^{*})\Big\}^2.
    \end{equation}
\end{thm}

As mentioned earlier, these are the same size and mixing conditions
used by \citet{Jon:97}.
Note that de Jong does allow a little bit more flexibility in the
conditions on the array $\{c_{nt}\}$ \citep[see also][]{Dav:93};
essentially, he allows for a single set of blocks with the
maximal $\{c_{nt}\}$ over each block well-behaved, while this
paper requires this condition to hold for every possible partition of
blocks.  This additional restriction is required because the
Stationary Bootstrap will select the blocks randomly and is similar to
\citepos{JoD:00b} requirement for the \fclt. Similarly, our assumption
on the dispersion of the individual means, $(\mu_{nt} - \bar\mu_n)^2$, is
slightly stronger than \citepos{GoW:02} and \citepos{GoJ:03} to
accommodate larger block sizes.

Theorem~\ref{main-bootstrap-clt} relies on a general insight about the
variance of the sample mean under the bootstrap-induced
distribution. It is well-known that a key step in proving the \clt\
for arbitrary dependent processes is demonstrating that the squared
elements converge to a positive and finite limit; i.e.\ if
$\{Z_{nj};~j=1,\dots,J_n\}$ is a representative stochastic array,
$\sum_j Z_{nj}^2 \to^p \sigma^2$ is an important necessary condition
for $\sum_j Z_{nj} \to^d N(0,\sigma^2)$.  (See Section 3.2 of
\citealp{HaH:80}, for further discussion.)  For martingale difference
arrays, each $Z_{nj}$ is one of the original random variables $X_{nt}$
(typically normalized by $1/\sqrt{n}$), but for other forms of
dependence (\ned\ or mixingale arrays, for example, as in
\citealp{Jon:97}) each $Z_{nj}$ is a contiguous block of the original
random variables,
\begin{equation*}
  Z_{nj} = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n +1}^{j M_n} (X_{nt} - \mu_{nt}),
\end{equation*}
that adds up to the original summation, plus potentially a negligible
residual, so $\sum_j Z_{nj}= (1/\sqrt{n}) \sum_t (X_{nt} - \mu_{nt}) +
o_p(1)$. In \citet{Jon:97}, for example, the \clt\ for mixingale
arrays assumes that there exists a sequence of blocks such that $\sum_j
Z_{nj}^2$ converges i.p., and the \ned\ \clt\ establishes conditions
under which such a $Z_{nj}$ exists.

Our insight is that the expectation of squared blocks of the bootstrap
process can be expressed as a sequence of \emph{contiguous} blocks of
the original process, so the arguments that establish convergence of
the original squared blocks can be applied with only minor changes to
the bootstrapped blocks. Consider the Moving Blocks Bootstrap,%
\footnote{To make this presentation as simple as possible, assume for
  now that $n = M_n J_n$ exactly.} %
for example, and let
\begin{equation*}
  Z_{nj}^* = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n + 1}^{j M_n} (X_{nt}^* - \bar X_n).
\end{equation*}
Then, conditional on the data, the $Z_{nj}^{*2}$ are independent can
be expected to obey an \lln, so %
$\sum_j (Z_{nj}^{*2} - \E^* Z_{nj}^{*2}) \to^p 0$ %
and the CLT for the bootstrapped array requires $\sum_j \E^*
Z_{nj}^{*2}$ to converge to a positive and finite limit.  But, since
$\E^*$ only averages over the starting point of each block, we have
\begin{align*}
  \E^* Z_{nj}^{*2} &=
  \tfrac{1}{n} \sum_{\tau = 0}^{n-M_n}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= \tau + 1}^{\tau + M_n} (X_{nt} - \bar X_n) \Big)^2 \\
  &= \tfrac{1}{n} \sum_{\tau_0 = 0}^{M_n-1}
  \sum_{j=1}^{J_n-1}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= (j - 1) M_n + \tau_0 + 1}^{j M_n + \tau_0} (X_{nt} - \bar X_n) \Big)^2
\end{align*}
after grouping blocks separated by $M_n$ periods. For each $\tau_0$,
the summation can be expected to converge in probability through the
same arguments that were used to establish the \clt\ for the original
array.%
\footnote{Some of the details of the argument will typically need to
  change, since the original \clt\ only requires convergence for
  $\tau_0 = 0$, but these details are often incidental to the original
  argument.}%
\footnote{Lemmas~\ref{bootstrap-variance-convergence-lemma} and
  \ref{bootstrapped-zsq-uniform-integrability-lemma} are particularly
  strong demonstrations of this argument.} %
A similar representation is available for the Circular and Stationary
bootstraps.

In short, the basic approach that we use to prove Theorem~\ref{main-bootstrap-clt}
is based on a fundamental connection between the second moments of the
bootstrap process and the sum of squared blocks of the original
array. Even though the details of our proof rely on specific
techniques for \ned\ arrays, this connection implies that block
bootstraps are typically consistent when the original dependent array
obeys the \clt, and the connection should be useful for proving
consistency of the bootstrap under other dependence conditions.

For some uses, a \clt\ is inadequate and an \fclt\ is necessary.
Theorem~\ref{main-bootstrap-clt} can be extended to give an \fclt\
(using arguments from \citealp{JoD:00b}). We show in
Theorem~\ref{main-bootstrap-invariance-theorem} that the partial sum
of the bootstrapped process obeys an \fclt\ and can be used to derive
critical values for other test statistics under the same assumptions
as Theorem~\ref{main-bootstrap-clt}.

\begin{thm}\label{main-bootstrap-invariance-theorem}
  Suppose that the conditions of Theorem~\ref{main-bootstrap-clt}
  hold, let $W$ be standard Brownian Motion, and define
  \begin{equation}
    \label{eq:6}
    W_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \sigma_{n}^{*}
  \end{equation}
  and
  \begin{equation}
    \label{eq:7}
    \hat{W}_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \hat\sigma_n^{*}.
  \end{equation}
  Then $\pr^{*}[\dist(W_n^{*}, W) > \delta] \to 0$ i.p.\ and
  $\pr^{*}[\dist(\hat{W}_n^{*}, W) > \delta] \to 0$ i.p.\ for any
  positive $\delta$ and distance $\dist$ that metricizes weak
  convergence.
\end{thm}

Obviously, Theorem~\ref{main-bootstrap-invariance-theorem}, has
natural corollaries that allow it to be used to approximate the
behavior of partial sums of the original series.  But applying this
result requires stronger assumptions than
Theorem~\ref{main-bootstrap-clt}. The bootstrapped processes are
(essentially) homoskedastic, so they are unable to match patterns of
heteroskedasticity in the original series and we can not estimate the
covariance process of the original partial sum in the bootstrap. This
necessitates a separate consistent estimator of the covariance process, which
we did not need for the sample mean.  Other methods, such as the Local
Block Bootstrap \citep{PaP:02,DPP:03}, may be able to capture this
additional heterogeneity with the bootstrap alone, but we do not
pursue that possibility further.
